Replacing standard galaxy profiles with \\ mixtures of Gaussians

### 2013-02-16 documents/mixtures/response_to_referee.txt

   This paper gives a particularly lucid, concise and clear
   description of a relatively old idea. As such, it represents a
   significant contribution to the literature as a pedagogical
   reference. It should be published but see 4. below.

We agree with the referee that the idea of using Gaussians to fit
galaxies is old but we don't agree with any implication that the
method or applied math results shown are old.  All references we can
find in the literature show *how* to do mixture-of-Gaussian fitting in
real imaging, but don't give mixture-of-Gaussian approximations to
known profiles.  This paper, in that sense, is more of an "applied
math meets astronomy" paper than the previous work.  We will return to
this below, but we *did* make changes to the text in response to this
general concern of the referee: We added sentences in the introductory
paragraphs and at the end of the paper to emphasize the difficulty of
rendering the profiles without these approximations.

To be more specific: we added some sentences to emphasize that the
currently used methods for rendering profiles are effectively mixtures
of delta functions (the profile is sampled at a finite set of
locations) and rendering requires as many PSF renderings as there are
sample points.  This is a huge computation.

We also added an introductory paragraph that points out that the code
complexity required to do rendering correctly is large and this leads
to engineering overheads in testing and correctness that go way beyond
the simple numerical speed losses.  We alsoo bring up the point that
conventional numerical integration methods effectively treat the
profiles as very complex mixtures of kernels.

[ should we also add a late paragraph re-iterating and showing a
comparison? ]

@@

   Page 6, first line: "The half-light inverse radius parameters
   \alpha^{exp} and \ alpha^{dev} -- and the softening and cutoff
   radius parameters -- are taken from the SDSS codebase." I think you
   meant to write \alpha^{lux} and \alpha^{luv} here.

Absolutely; fixed.

   Page 6, paragraph 2: "Although there is an analytic result for the
   dev profile, numerical integration of the concentrated profiles dev
   and luv to determine total fluxes can be challenging."  Why are you
   flagging the numerical integration of the dev profile as a
   challenge given that there is an analytic expression for it?

Indeed.  Adjusted that paragraph slightly and added a subsequent
paragraph explaining the implications for integrals over finite
domains.

@@

   Page 6, paragraph 2: "K-L divergence" --> "Kullback-Leibler
   divergence"

fixed.

   Page 6 - Definition of "best": The authors should really have been
   "bolder" by insisting on the use of an information-theoretic
   measure here rather than the "traditional" chi-squared. The MoGs
   are a natural fit to an image entropy measure because maximizing
   entropy, say, would naturally lead to minimizing the number of
   components in a MoG. Chi-squared is also biased at low
   signal-to-noise ratios.  Please consider.

Chi-squared would definitely be biased if what we were fitting in
astronomical images were distributions of photon arrivals with no
background or read noise.  The presence of background (sky photons)
and read noise makes chi-squared a better statistic for optimization
than, say, the Poisson likelihood.  However, the *precise* measure you
would use in *principle* depends (weakly) on the properties of the
data you have.

[ make text change here ]

As for the information-theoretic things mentioned by the referee: Yes,
in principle things like AIC and BIC and Bayes integral etc measures
for "how many components to use" would be very useful.  Unfortunately,
unlike with the decision about chi-squared vs Poisson likelihood, the
value of the AIC or BIC does depend very very sensitively on the noise
model (object, sky, and read noise levels) in every pixel of every
image.  That is, someone doing this correctly would use a higher K
expansion for the brighter galaxies and a lower K expansion for the
fainter galaxies.  That is very relevant and important...

[ we must add text for this ]

@@

   Page 6: Why is "badness" computed in a one-dimensional numerical
   integral? For the sake of computational speed? What happens when
   dealing with heavily blended objects? Where is the center for the
   1D integral then? Please explain. A general explanation of how
   object deblending is performed with MoGs would be welcome.

All we are trying to do in this paper is come up with *replacements*
for the exponential and deVaucouleurs profiles and variants.  We are
not actually fitting real galaxy images.  Indeed that requires
deblending and so on.  Literally, all we are doing is replacing the
exp{r^{1/n}) with a sum of Gaussians.  We are not fitting real scenes.

In the Tractor papers which are forthcoming, we *are* solving this
deblending problem, but that is a separate matter, involving image
likelihoods, noise models, and priors (and utilities, perhaps) on
parameters, so it is way beyond the scope of this paper.

In this paper *all* we are doing is the "applied math" problem of
replacing the analytic profiles with MoG approximations.

[ we must make a text change here in response to this issue ]

@@

   Page 7, paragraph 2: The statement "These -- M^lux = 6 and M^luv =
   8 -- are good compromises between mixture complexity and quality of
   fit" warrants more explanation. Everytime I read about using MoGs
   for galaxy fitting, I have the same reaction: it needs a large
   number of highly redundant to model something a single exponential
   profile. Fans of the eigenimage decompositions are probably
   appalled!  Granted the mathematical simplicity of MoGs can make up
   for this by enabling faster fitting times, but it is definitely an
   un-appealing aspect of the method.  It would be helpful to comment
   on this to counter a reaction that I think may be shared by
   others. More importantly, I do not quite understand how you decided
   on M^lux = 6 and M^luv = 8. If I look at Table 2, the badness for
   lux has clearly converged and there is no gain in going higher than
   M=6. I cannot say the same thing for the luv MoG. Badness goes down
   by a factor of 16 going from M=6 to M=8 and then goes down by a
   factor of 5 going from M=8 to M=10. How can you claim that your
   badness has "converged" for M=8 then? A quantitative threshold and
   its rationale are clearly needed here to justify the "good
   compromises" here. Please be more quantitative or clarify if I
   misunderstood something.

[ response ; what changes ]

   Equation 19: I did not quite understand your distinction between
   the \Delta X vector here and the x vector in equation 2. Please
   clarify.

[ response; what changes ]

[ and search email for other issues arising / suggestions / etc... ]

### 2013-02-17 documents/mixtures/mixture_models.tex
why Delta x and not x

The value of these MoG approximations comes when they are to be
convolved with a PSF (usually in fact a pixel-convolved PSF) that is
itself also represented as a MoG.  In this scenario, the PSF
$\psi(\Delta\pos)$---which is thought of as a function of focal-plane
displacement $\Delta\pos$ away from, say, a true stellar position---is
represented as a $K$-Gaussian MoG
\begin{eqnarray}\displaystyle
\psi(\Delta\pos) &=& \sum_{k=1}^K p_k\,\normal(\Delta\pos\given\mean_k,\var_k)
\\
1 &=& \sum_{k=1}^K p_k
\quad ,
\end{eqnarray}
where the means $\mean_k$ are \emph{not} required to vanish because
the PSF can have arbitrarily non-trivial structure (think speckles and
the like) and the variances $\var_k$ will not in general be
proportional to the identity or even diagonal because the PSF will not
in general be round.  An example that illustrates the use of this PSF
is the following: A star of flux $S_s$ at focal-plane position
$\pos_s$ will lead to an image (PSF-convolved intensity map) of the
form
\begin{eqnarray}\displaystyle
I(\pos\given\star,S_s,\pos_s) &=& \sum_{k=1}^K S_s\,p_k\,\normal(\pos\given\pos_s+\mean_k,\var_k)
\quad .
\end{eqnarray}
That is, when the PSF is represented as a MoG, any image of a
star---or indeed any image of any set of stars---is also represented
as a MoG.

>>>>

The value of these MoG approximations comes when they are to be
convolved with a PSF (usually in fact a pixel-convolved PSF) that is
itself also represented as a MoG.  In this scenario, the PSF
$\psi(\Delta\pos)$---which is thought of as a function of focal-plane
displacement $\Delta\pos$ away from, say, a true stellar position---is
represented as a $K$-Gaussian MoG
\begin{eqnarray}\displaystyle
\psi(\Delta\pos) &=& \sum_{k=1}^K p_k\,\normal(\Delta\pos\given\mean_k,\var_k)
\\
1 &=& \sum_{k=1}^K p_k
\quad ,
\end{eqnarray}
where the means $\mean_k$ are \emph{not} required to vanish because
the PSF can have arbitrarily non-trivial structure (think speckles and
the like) and the variances $\var_k$ will not in general be
proportional to the identity or even diagonal because the PSF will not
in general be round.  The reason we use $\Delta\pos$ in the argument
of $\psi(\cdot)$---and not $\pos$---is that the PSF describes the
probability distribution for two-dimensional \emph{displacements} of
detected photons from true source positions.  It is equivalent to say
that the PSF is a function that is only used to \emph{convolve} (or
correlate) two-dimensional models; it is never used as an image model
\emph{itself}.  The PSF is only used in the context of convolving some
high-resolution model, which might be as complex as a multi-component
galaxy, or might be as simple as a delta function, in the case of a
star.

An example that illustrates the use of this PSF is the following: A
star of flux $S_s$ at focal-plane position $\pos_s$ will lead to an
image (PSF-convolved intensity map) of the form
\begin{eqnarray}\displaystyle
I(\pos\given\star,S_s,\pos_s) &=& \sum_{k=1}^K S_s\,p_k\,\normal(\pos\given\pos_s+\mean_k,\var_k)
\quad .
\end{eqnarray}
That is, when the PSF is represented as a MoG, any image of a
star---or indeed any image of any set of stars---is also represented
as a MoG.
